My title

### 2024-11-25 paper/paper.qmd
add model

library(tidyverse)
library(ggplot2)
library(dplyr)
library(here)
library(reshape2)
library(scales)
library(arrow)
library(knitr)
library(kableExtra)
```

@@

Here we briefly describe the multiple linear regression model used to investigate these relationships. The model captures both linear and nonlinear effects (e.g., through the quadratic term for age), and all categorical variables are treated as factors to account for group-level differences. Background details, including diagnostics and model assumptions, are included in.

@@

Multiple regression models rely on several assumptions: linearity, meaning a linear relationship exists between independent and dependent variables; homoskedasticity, requiring constant error variance across predictors; independence of errors, meaning residuals are uncorrelated; normality, where residuals follow a normal distribution; and independence of independent variables, ensuring no significant multicollinearity.

@@

@@

@@

>>>>

library(tidyverse)
library(ggplot2)
library(dplyr)
library(here)
library(reshape2)
library(scales)
library(arrow)
library(knitr)
library(kableExtra)
library(car) 
library(broom) 
```

@@

Here we briefly describe the multiple linear regression model used to investigate these relationships. The model captures both linear and nonlinear effects (e.g., through the quadratic term for age), and all categorical variables are treated as factors to account for group-level differences.

@@

Multiple regression models rely on several assumptions: linearity, meaning a linear relationship exists between independent and dependent variables; homoskedasticity, requiring constant error variance across predictors; independence of errors, meaning residuals are uncorrelated; normality, where residuals follow a normal distribution; and independence of independent variables, ensuring no significant multicollinearity. These assumptions are discussed in @sec-model-details and evaluated using the diagnostic plots shown in @fig-check.

@@

The evaluation of the final model's performance was conducted on both training and testing datasets. As shown in tbl-test, the RMSE and MAE values are slightly higher for the testing data compared to the training data (0.786 vs. 0.773 for RMSE, and 0.568 vs. 0.543 for MAE), reflecting a modest decline in predictive accuracy when applied to unseen data. Similarly, the $R^2$ values are 0.488 for the training set and 0.468 for the testing set, indicating that the model captures a moderate proportion of the variability in the log-transformed income. Additionally, all variables included in the final model were statistically significant, with $p$-values below 0.05, further supporting the robustness of the selected predictors. Overall, the final model achieves a practical balance between complexity, interpretability, and predictive performance.

@@

@tbl-modelresults presents the effects of weekly working hours, age, region, education level, gender, and racial group on log income, along with their 95% confidence intervals. The estimated coefficients indicate varying degrees of influence among these factors.

Education level has the strongest impact on income. Compared to those with a high school education (the reference group), individuals with education beyond a bachelor’s degree have a coefficient of 1.084, showing a significant increase in income. Similarly, those with a bachelor’s degree and some college education have coefficients of 0.8713 and 0.5094, respectively, suggesting higher education correlates with higher income.

Age and its quadratic term reveal a nonlinear relationship. The positive coefficient for age (0.0985) indicates income increases with age. However, the negative coefficient for the quadratic term (-0.0010) shows this growth slows over time, eventually declining. This pattern reflects typical career income trajectories, with rapid growth early on, stability in middle age, and potential decreases later in life.

Gender also plays a significant role. Male income is significantly higher than female income, with a coefficient of 0.2632, even after controlling for other factors.

The effect of region varies. Compared to the Midwest (reference group), incomes in the West and Northeast are higher, with coefficients of 0.1560 and 0.1287. The coefficient for the South is -0.0192, close to zero, indicating little difference. This may relate to regional economic conditions and policies.

Weekly working hours show a modest effect on income, with a coefficient of 0.040. While longer hours increase income, the impact is relatively small, potentially due to wage structures and overtime policies.

The effects of racial group are more complex. Compared to Asian individuals (reference group), income for Black and "Other" racial groups is notably lower, with coefficients of -0.1014 and -0.1290. The coefficient for White individuals is nearly zero (-0.0007), indicating no significant difference. These patterns reflect complexities in racial income disparities.

In summary, education level has the largest influence on income, followed by age and gender. Region and racial group also play roles, while the effect of weekly working hours is smaller. These results provide a clear picture of how socioeconomic factors affect income and offer important context for further research and policymaking.

@@

```{r}
#| echo: false
#| warning: false
#| label: fig-check
#| fig-cap: "Model Assumption Checks"
#| fig-subcap: [
#| "Residuals vs Fitted", "Normal Q-Q Plot", 
#| "Scale-Location", "Residuals vs Leverage"
#| ]
#| layout-ncol: 2

# Extract residuals and fitted values from the model
model_data <- augment(final_model)

# Check linearity assumption: Residuals vs Fitted Plot
residuals_vs_fitted <- ggplot(model_data, aes(.fitted, .resid)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  labs(x = "Fitted values", y = "Residuals") +
  theme_minimal()

# Check normality of residuals: Normal Q-Q Plot
qq_plot <- ggplot(model_data, aes(sample = .std.resid)) +
  stat_qq() +
  stat_qq_line(color = "red") +
  labs(x = "Theoretical Quantiles", y = "Standardized Residuals") +
  theme_minimal()

# Check homoscedasticity: Scale-Location Plot
scale_location <- ggplot(model_data, aes(.fitted, sqrt(abs(.std.resid)))) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  labs(x = "Fitted values", y = "Sqrt(Standardized Residuals)") +
  theme_minimal()

# Check for outliers and leverage points: Residuals vs Leverage
residuals_leverage <- ggplot(model_data, aes(.hat, .std.resid)) +
  geom_point(alpha = 0.6) +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  labs(x = "Leverage", y = "Standardized Residuals") +
  theme_minimal()

residuals_vs_fitted
qq_plot
scale_location
residuals_leverage

```
The model assumptions were checked using @fig-check. First, the Residuals vs Fitted plot tests the linearity assumption, showing that residuals are randomly scattered around the fitted values without any clear pattern, indicating that the linear relationship is valid. Second, the Normal Q-Q plot checks for the normality of residuals. Most points align closely with the diagonal line, suggesting the residuals are approximately normally distributed, with only slight deviations at the tails. Third, the Scale-Location plot evaluates homoskedasticity, and the red line remains mostly flat, suggesting that the variance of the residuals is fairly consistent. Finally, the Residuals vs Leverage plot identifies potential outliers or high-leverage points, indirectly testing the independence of errors and the independence of variables. Most points have low leverage, with only a few high-leverage points that may need attention. Overall, the assumptions are largely met, though there are some limitations that could be addressed with further refinements in the analysis.
